---
title: "Calling R functions from Fortran"
date: '`r Sys.Date()`'
author: "Balasubramanian Narasimhan"
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{frcall}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction

The _Writing R extensions_ manual actually provides almost all the
details for calling R functions from Fortran. This package is a
concrete realization of those instructions. There are, however, a few
details to note in the interest of portability.

- The only structures passed to and fro the R function from Fortran
  are integer and double vectors or matrices.
- Strings are problematic and best avoided.
- A bit of C glue is needed to make the call to R. 

## Is it even needed?

There are times when you would like to allow a user-specified
computation to be used in your awesome Fortran routine. In fact, some
of the fastest algorithms are implemented in Fortran by Stanford
colleagues of mine and they've asked me for such a facility.


## Example

Suppose we wish to call the following R function from Fortran in a
package, say `ftest`. This is where it is most useful anyway.

```{r}
#' Compute a discrepancy measure between y and z weighted by w
#' @param y the first vector
#' @param z the second vector
#' @param w the weights
#' @return a scalar discrepancy measure
discrepancy  <- function(y, z, w) {
    w  <- w/sum(w)
    sum( w * (y - z)^2 )
}
```

An example call:

```{r}
discrepancy(y = 1:10, z = 11:20, w = rep(1, 10))
```

## Calling from Fortran

The steps are as follows, with some being specific to the R function
being called. The calling sequence involves two calls that go together
in sequence: saving the R function in a place that Fortran can look
for it and calling a Fortran function to actually execute the call.

1. Write a C function, say `src/rcall.c` to call the R function as
   follows. What follows below is the implementation of the
   `discrepancy()` function above. You will have to tailor this to
   each R function you want to call. Note that this requires knowledge
   of C/C++ interface in R, discussed in the _Writing R Extensions_
   manual. The comments below are meant to help.

```{c, eval = FALSE}
/* Start of section not to be modified */

#include <R.h>
#include <Rinternals.h>

static SEXP rfunc;

/* Store the R function */
SEXP store_rfun(SEXP rfun) {
  rfunc = rfun;
  return(R_NilValue);
}

/* End of section not to be modified */

void F77_NAME(rfcall)(int *n, double *y, double *z, double *w, double *result) {

  /* 
	 Ensure you modify the arguments above first for your function,
     then modify stuff below to match that.  Note that the length of
     the y, z, w arrays has to be passed, and the result is stored
     into a variable that Fortran provides `result`.
  */

  int num_protected = 0;   /* The number of args protected on stack */
  SEXP ry, rz, rw;  /* The R structures we will build below. */

  /* Allocate space for the vectors we wish to construct */
  PROTECT(ry = allocVector(REALSXP, *n));
  num_protected++;
  PROTECT(rz = allocVector(REALSXP, *n));
  num_protected++;
  PROTECT(rw = allocVector(REALSXP, *n));
  num_protected++;

  /* Now copy the values from Fortran arrays into the R arrays */
  double *yvec = REAL(ry); double *zvec = REAL(rz); double *wvec = REAL(rw);
  for (int i = 0; i < *n; i++) {
    yvec[i] = y[i]; zvec[i] = z[i]; wvec[i] = w[i];
  }

  /* Standard stuff: set up environment and evaluate R function */
  SEXP rho = R_GetCurrentEnv();
  SEXP call = PROTECT(LCONS(rfunc, LCONS(ry, LCONS(rz, LCONS(rw, R_NilValue)))));
  num_protected++;
  SEXP r_result = R_forceAndCall(call, 3, rho);

  /* Examine result length */
  R_len_t len = length(r_result);

  /* Throw error if result is more than one; might bomb anyway */
  if (len > 1) {
    error("R discrepancy function result length > 1");
  }

  *result = REAL(r_result)[0];

  UNPROTECT(num_protected);
}
```

2. Write a Fortran function to call the C function in step 1.

```{fortran, eval = FALSE}
c$$$  Assumes the r function has been set using ftest::save_rfun
      subroutine fcallr(n, y, z, w, res)
      integer n
      double precision y(n), z(n), w(n), res
      call rfcall(n, y, z, w, res)
      return
      end
```

3. Write an R function to save the R function where Fortran can find
   it. This is problem invariant.

```{r, eval = FALSE}
#' Save the function f for calling from fortran
#' @param f the R function to be called using `.Fortran`
#' @useDynLib ftest
#' @export save_rfunc
save_rfunc  <- function(f) {
    .Call("store_rfun", f, PACKAGE = "ftest")
    invisible(TRUE)
}
```

4. Write the Fortran interface using `.Fortran()` call.

```{r}
#' Compute the discrepancy using `r_fun` using`fcallr`
#' @para r_fun a function of three parameters 
r_from_f  <- function(r_fun, y, z, w) {
    ## save the R function to call in C ; fcallr knows to look for it there.
    ftest::save_rfunc(r_fun)
    u <- .Fortran("fcallr",
                  length(y),
                  as.double(y),
                  as.double(z),
                  as.double(w),
                  result = as.double(0), PACKAGE = "ftest")
    u$result
}
```

## Test Run

An actual run

```{r}
y = 1:10; z = 11:20; w = rep(1, 10);
cat(sprintf("Direct R function call: %f\n", discrepancy(y = y, z = z, w = w)))
cat(sprintf("Calling R from Fortran: %f\n", r_from_f(discrepancy, y, z, w)))
```












